High-Order Accurate Finite Difference Schemes \\ Via Optimal Time Step Selection

### 2008-06-09 papers/OTSPDE/OTSPDE.tex
OTSPDE:  final submission revision of OTSPDE paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{Using Optimal Time Step Selection to Boost the Accuracy of
       Finite Difference Methods for a Class of 
       Time Dependent Partial Differential Equations}

@@

\begin{abstract}
In this article, we present a novel technique for obtaining high-order
accurate numerical solutions to time dependent partial differential 
equations using only formally low-order finite difference stencils and
time integration schemes.  We achieve high-order accuracy in a simple
manner by using an optimally chosen time step (and possibly adding a few 
correction terms to the finite difference scheme).  Through straightforward 
numerical analysis arguments, we explain the origin of the boost in accuracy 
and estimate the computational cost of the high-order methods obtained via 
optimal time step (OTS) selection.  We demonstrate the utility of OTS 
selection on a wide array of classical PDEs in one and more space dimensions
and find that it is applicable to linear and semilinear PDEs on both regular 
and irregular domains. 
\end{abstract}

@@

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{KEVIN T. CHU}
         {HIGH-ORDER FINITE DIFF SCHEMES VIA OPTIMAL TIME STEP SELECTION}

@@

There are many high-order accurate methods that are simple combinations or 
modifications of low-order schemes.  A few well-known examples include 
Richardson extrapolation~\cite{richardson_1927, heath_book}, the modified 
nine-point formula for the 2D Poisson equation~\cite{iserles_book}, and the 
Lax-Wendroff method for the advection equation~\cite{leveque_book_1992, 
leveque_book_2002, gko_book}.  Cancellation of higher-order errors (whether 
fortuitous or intentional) is a key principle underlying the superior 
accuracy of these methods.

More commonly, high-order accurate numerical methods are constructed by 
requiring that the discretization error possesses the desired formal order of 
accuracy for \emph{all} of the independent variables in the PDE.  While this 
constraint is certainly sufficient for high-order accuracy, it is by no means 
necessary.  This conventional approach for constructing numerical methods 
has its roots in the classical numerical analysis of PDEs, which tends to 
analyze the discretization of the terms in the PDE individually.  Within this 
framework, it is natural to seek high-order accuracy by ensuring that the 
numerical scheme is simultaneously high-order accurate for all independent 
variables.  Unfortunately, this common approach for developing and analyzing 
numerical methods is often suboptimal because it neglects the potential for 
achieving high-order accuracy using only simple, low-order schemes.

Perhaps the greatest weakness of the standard method for constructing 
high-order numerical methods is the use of ``big-O" notation, which hides 
the details of the discretization error.  The information lost in the 
``big-O" notation eliminates the possibility of reducing numerical error by 
taking advantage of relationships between the terms in the error.  The 
modified nine-point formula for the Poisson equation~\cite{iserles_book} 
and the Lax-Wendroff scheme for the advection equation~\cite{leveque_book_1992,
leveque_book_2002, gko_book} demonstrate the power of explicitly retaining 
higher-order terms in the discretization error when analyzing numerical 
schemes.  For both of these methods, examination of higher-order terms allows 
low-order methods to be transformed into higher order methods with only minor 
modifications to the numerical scheme.

The usual approach for developing numerical methods has another significant
shortcoming -- it places no value on the optimization of numerical parameters 
(\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
somewhat to blame for this situation because it gives the false impression 
that the error depends on each numerical parameter independently.  Aside
from constraints needed for numerical stability, ``big-O" notation makes 
it appear as though any combination of numerical parameters is equally good 
as long as they are all ``small enough.''  For some problems, however, 
optimizing the numerical parameters can significantly increase the accuracy
of the numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed 
that choosing an appropriate value for $\theta$ as a function of the time 
step in the theta-method\footnote{In~\cite{zhao_2006} refers to $\theta$ as 
the weight factor and the theta-method as the single-step trapezoidal 
method.}~\cite{iserles_book, gko_book} can decrease the error of finite 
element solutions to the diffusion equation.

@@

Using the PDE being solved to reduce discretization error has a long 
tradition in numerical analysis.  Indeed, this idea underlies both the 
modified nine-point formula for the Poisson equation~\cite{iserles_book} and 
the Lax-Wendroff method for the linear advection 
equation~\cite{leveque_book_1992, leveque_book_2002, gko_book}.  
In the case of the modified nine-point formula for the Poisson equation, a 
simple analysis of the nine-point stencil for the Laplacian on a uniform grid 
shows that it is only second-order, which might lead to the conclusion that 
the nine-point stencil has no advantage over the standard five-point stencil.  
However, a more careful analysis of the discretization error in light of the 
PDE reveals that the leading-order error, which is proportional to the 
bilaplacian of the solution~\cite{iserles_book, patra_2005}, is directly 
related to the Laplacian of the source term in the Poisson equation.  Thus, 
the numerical scheme is easily boosted to fourth-order by simply adding a 
correction term that cancels out the leading-order error.   
Similarly, use of the PDE to eliminate higher-order discretization errors also 
plays a critical role in the Lax-Wendroff method because it allows the second 
derivative of the solution in time to be written in terms of the second 
derivative in space.  This observation transforms an unstable, first-order 
scheme based on a central difference approximation of 
$\partial/\partial x$ into a stable, second-order 
scheme~\cite{leveque_book_1992, leveque_book_2002}.

Optimal time step selection as a technique for improving the overall accuracy 
of numerical schemes for time dependent PDEs appears to be a novel idea.
Typically, the primary concern when choosing a time step is numerical 
stability with accuracy being a secondary consideration.  The fact that a 
clever choice of time step can be used to dramatically reduce the error is 
rarely considered.  In general, little research seems to have been done in
the area of optimizing the parameters of numerical methods (\eg time step and 
grid spacing) to improve accuracy.  To the best of the author's knowledge, 
Zhao's method for choosing an optimal value of $\theta$ in the theta-method is 
the only work that attempts to improve accuracy by modifying numerical 
parameters.  While he focuses solely on the optimal choice of $\theta$ for a 
fixed time step, his approach can be viewed as a form of optimal time step 
selection by reversing the roles of $\theta$ and the time step.

It is worth spending a few moments to draw the distinction between optimal 
time step selection and traditional adaptive time stepping 
techniques~\cite{iserles_book,??} that are used in the context of the method 
of lines~\cite{iserles_book, gko_book}.  On the surface, optimal time step 
selection may appear to be just a crude version of adaptive time step control.
However, the two methods are actually very different.  While both methods seek 
to reduce numerical error by controlling the time step, the numerical errors 
they control are fundamentally different.  Adaptive time stepping can only 
be used to reduce temporal discretization errors because it controls the 
errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Thus, optimal time 
step selection can potentially produce superior results compared with 
traditional adaptive time stepping approaches for integrating the 
semi-discrete equations associated with time dependent PDEs.  Another 
important distinction between the two techniques is that optimal time step 
selection uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

@@

The remainder of this paper is organized as follows.  
In Section~\ref{sec:optimal_time_step_selection}, we begin by 
discussing the technique of optimal time step selection in detail and 
presenting the analysis that shows how it boosts the order of accuracy 
for formally low-order numerical schemes.  In 
Section~\ref{sec:applications_1d}, we demonstrate the general utility of 
optimal time selection to problems in one space dimension by applying it to 
finite difference schemes for several classical linear and nonlinear PDEs. 
In Section~\ref{sec:applications_multidim}, we use optimal time step selection 
to improve the accuracy of simple finite difference schemes for the linear 
advection equation and diffusion equation in multiple space dimensions.  
To illustrate the ease with which high-order accuracy can be achieved for
problems on arbitrary domains, the 2D diffusion equation is solved on both 
the regular and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size and order of accuracy.  
Finally, in Section~\ref{sec:summary}, we provide some concluding remarks.

\section{\label{sec:optimal_time_step_selection} 
         Optimal Time Step (OTS) Selection}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather it is a way to enhance to the performance
of existing finite difference schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  The two basic ideas underlying 
optimal time step selection are (1) the PDE provides valuable insight into the 
discretization errors for finite difference schemes and (2) a judicious choice 
of time step can be used to eliminate the leading-order (and possibly 
higher-order) terms in the error.  In this section we present and analyze the 
technique of optimal time step selection for finite difference schemes 
constructed to solve evolution equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important in 
many problems of applied mathematics, restricting our attention to problems 
of this form is not a serious limitation.

@@

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the finite difference approximation of the spatial 
derivative operator acting on ${\mathbf u}$.  The second term in the local 
error arises directly from temporal discretization errors associated with the 
numerical scheme used to integrate the system of ODEs in time.  The first 
term arises because $F$ is only an approximation of the spatial derivative 
operators in the PDE.  As a result, the system of 
ODEs~(\ref{eq:method_of_lines}) has an error due to the spatial discretization 
at \emph{every} point in time.  For a spatial discretization error of order 
$r$, the error in (\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the 
error accumulated during each time step due to spatial discretization error is 
$O(\dt \dx^r)$ regardless of how the system of ODEs (\ref{eq:method_of_lines})
is solved numerically.  Because the method of lines is so frequently used to 
construct finite difference approximations for time dependent PDEs, we will 
focus solely on finite difference schemes of this type for the remainder of 
the paper to keep the discussion simple.  It is straightforward to extend the 
results we obtain to more general finite difference schemes.

To analyze the accuracy of the finite difference scheme 
for~(\ref{eq:model_PDE}) when the optimal time step is used, we examine the 
higher-order terms in the discretization error following the conventional 
approach (\ie ``big-O'' notation).  Elimination of the leading terms in the 
discretization error leaves a local error of the form 
$O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ and $(q+1) > (s+1)$ are the orders 
of the spatial and temporal discretization errors, respectively, after the 
leading-order terms in the error have been removed.  As for the leading-order 
error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
error is a direct consequence of the method of lines approach to discretizing 
the PDE.

@@

% TABLE SHOWING COMPUTATIONAL COSTS DISCUSSED IN SECTION
% sec:computational_performance.  PLACED HERE TO SPREAD TABLES OUT
\begin{table}[tb]
\caption{\label{tab:comp_perf_vs_err} 
   Computational cost of various finite difference schemes for the 1D 
   diffusion equation as a function of the global numerical error $e$.
   For all of the schemes, the standard second-order central difference 
   stencil is used to discretize the Laplacian, and the numerical 
   solution is assumed to be computed over a fixed time interval.  
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|l|c|c|c|c|}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$\footnote{The 
   time step for the forward/backward Euler and Crank-Nicholson schemes 
   are $O(\dx^2)$ and $O(\dx)$, respectively.  Note that the 
   time step for the backward Euler scheme is chosen so that the temporal 
   error is subdominant to the spatial error.}
  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
    to be $O(1/\dx)$.} 
  & {\bf Compute Time}\footnote{The computation time is approximated as 
    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
    arise for the implicit methods.}  \\
  \hline 
  Forward Euler    & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Backward Euler   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-1} \right)$ \\
  Forward Euler with OTS  & $O\left( e^{1/4} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/4} \right)$ 
                   & $O\left( e^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{center}
\end{minipage}
\end{table}

@@

@@

\begin{table}[tb]
\caption{\label{tab:comp_perf_vs_dim}
   Computational cost of the forward Euler scheme with optimal time step and 
   the Crank-Nicholson method for solving the diffusion equation as a function 
   of the global numerical error $e$.
   For both schemes, second-order, isotropic stencils are used to discretize 
   the Laplacian.  For a problem in $d$ space dimensions, the memory 
   requirement and computation time for each scheme are estimated to be 
   $O \left(1/\dx^d \right)$ and 
   $O\left( \left(1/\dx^d \right) (1/\dt) \right)$, respectively, which 
   assumes that we are using an efficient 
   $O(N) = O \left( 1/\dx^d \right)$ algorithm to invert the matrices that 
   arise for the Crank-Nicholson method.
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|c|c|}
  \hline
  & \multicolumn{2}{|c|}{{\bf Forward Euler with OTS}\footnote{For the forward 
    Euler scheme, the optimal time step is $\dto = \dx^2/6$.}}
  & \multicolumn{2}{|c|}{\bf Crank-Nicholson} \\
  \cline{2-3} \cline{4-5} 
    {\bf Dimensions} & {\bf Memory} & {\bf Compute Time} 
  & {\bf Memory} & {\bf Compute Time} \\
  \hline 
  $1$ & $O\left( e^{-1/4} \right)$ 
      & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ \\ 
  $2$ & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-3/2} \right)$ \\ 
  $3$ & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-5/4} \right)$ 
      & $O\left( e^{-3/2} \right)$ 
      & $O\left( e^{-2} \right)$ \\
  $d$ & $O\left( e^{-d/4} \right)$ 
      & $O\left( e^{-(d+2)/4} \right)$ 
      & $O\left( e^{-d/2} \right)$ 
      & $O\left( e^{-(d+1)/2} \right)$ \\ 
  \hline 
\end{tabular}
\end{center}
\end{minipage}
\end{table}

@@

\subsection{Boundary Conditions}
In our analysis and discussion thus far, we have completely ignored the issue 
of boundary conditions.  The choice of discretization for boundary conditions
can have a significant impact on the accuracy of the numerical solution.  In 
general, high-order discretizations of boundary conditions are required to 
ensure that errors introduced at the domain boundary do not pollute the 
numerical solution in the domain interior.  For Dirichlet boundaries, it
is sufficient to impose boundary conditions by using boundary (or ghost cell) 
values that have the same order of accuracy as the solution.  
More effort is required to ensure that Neumann boundary conditions do not
reduce the order of accuracy of the solution.  In general, Neumann boundary
conditions require that the accuracy of ghost cells slightly exceed the 
desired order of accuracy for the solution.  A detailed discussion of 
methods for imposing boundary conditions is beyond the scope of this paper.  
We refer interested readers to the thorough analysis presented
in~\cite{gko_book}.  To avoid complicating the discussion in 
Sections~\ref{sec:applications_1d} and~\ref{sec:applications_multidim}, 
Neumann boundary conditions are imposed by taking advantage of knowledge of 
the analytical solution.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
% Submission draft: 2008/06/08, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{High-Order Accurate Finite Difference Schemes \\
       Via Optimal Time Step Selection}

@@

\begin{abstract}
In this article, we present a novel technique for computing high-order
accurate numerical solutions to time dependent partial differential 
equations using only formally low-order finite difference stencils and
time integration schemes.  We achieve this high-order accuracy in a simple 
manner by optimally choosing the time step for the low-order finite 
difference scheme (and possibly adding a few correction terms).  Through 
straightforward numerical analysis arguments, we explain the origin of the 
boost in accuracy and estimate the computational cost of the high-order 
methods obtained via optimal time step (OTS) selection.  We demonstrate the 
utility of OTS selection on a wide array of classical PDEs in one and more 
space dimensions and find that it is applicable to linear and semilinear PDEs 
on both regular and irregular domains. 
\end{abstract}

@@

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{KEVIN T. CHU}
         {HIGH-ORDER FD SCHEMES VIA OPTIMAL TIME STEP SELECTION}

@@

There are many high-order accurate methods that are simple combinations or 
modifications of low-order schemes.  A few well-known examples include 
Richardson extrapolation~\cite{richardson_1927, heath_book}, the modified 
nine-point formula for the 2D Poisson equation~\cite{iserles_book}, and the 
Lax-Wendroff method for the advection equation~\cite{leveque_book_1992, 
leveque_book_2002, gko_book}.  Cancellation of higher-order errors is a key 
principle underlying the superior accuracy of these methods.

More commonly, high-order accurate numerical methods are constructed by 
requiring that the discretization error possesses the desired formal order of 
accuracy for \emph{all} of the independent variables in the PDE.  While this 
constraint is certainly sufficient for high-order accuracy, it is by no means 
necessary.  This conventional approach for constructing numerical methods 
has its roots in the classical numerical analysis of PDEs, which tends to 
analyze the discretization of the terms in the PDE individually.  Within this 
framework, it is natural to seek high-order accuracy by ensuring that the 
numerical scheme is simultaneously high-order accurate for all independent 
variables.  Unfortunately, this common approach for developing and analyzing 
numerical methods is often suboptimal because it neglects the potential for 
achieving high-order accuracy using simple low-order schemes.

Perhaps the greatest weakness of the standard method for constructing 
high-order numerical methods is the use of ``big-O" notation, which hides 
the details of the discretization error.  The information lost in the 
``big-O" notation eliminates the possibility of reducing numerical error by 
taking advantage of relationships between the terms in the error.  The 
modified nine-point formula for the Poisson equation~\cite{iserles_book} 
and the Lax-Wendroff scheme for the advection equation~\cite{leveque_book_1992,
leveque_book_2002, gko_book} demonstrate the power of explicitly retaining 
higher-order terms in the discretization error when analyzing numerical 
schemes.  For both of these methods, examination of higher-order terms allows 
low-order methods to be transformed into high-order methods with only minor 
modifications to the numerical scheme.

The usual approach for developing numerical methods has another significant
shortcoming -- it places no value on the optimization of numerical parameters 
(\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
somewhat to blame for this situation because it gives the impression 
that the error depends on each numerical parameter independently.  Aside
from constraints needed for numerical stability, ``big-O" notation makes 
it appear as though any combination of numerical parameters is equally good 
as long as they are all ``small enough.''  For some problems, however, 
optimizing the numerical parameters can significantly increase the accuracy
of the numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed 
that choosing an appropriate value for $\theta$ as a function of the time 
step in the theta-method\footnote{In~\cite{zhao_2006} refers to $\theta$ as 
the weight factor and the theta-method as the single-step trapezoidal 
method.}~\cite{iserles_book, gko_book} can decrease the error of finite 
element solutions to the diffusion equation.

@@

Using the PDE to reduce discretization error has a long 
tradition in numerical analysis.  Indeed, this idea underlies both the 
modified nine-point formula for the Poisson equation~\cite{iserles_book} and 
the Lax-Wendroff method for the linear advection 
equation~\cite{leveque_book_1992, leveque_book_2002, gko_book}.  
In the case of the modified nine-point formula for the Poisson equation, a 
simple analysis of the nine-point stencil for the Laplacian on a uniform grid 
shows that it is only second-order accurate, which might lead to the conclusion 
that the nine-point stencil has no advantage over the standard five-point 
stencil.  However, a more careful analysis of the discretization error in 
light of the PDE reveals that the leading-order error, which is proportional 
to the bilaplacian of the solution~\cite{iserles_book, patra_2005}, is directly 
related to the Laplacian of the source term in the Poisson equation.  Thus, 
the numerical scheme is easily boosted to fourth-order by simply adding a 
correction term that cancels out the leading-order error.   
Similarly, use of the PDE to eliminate higher-order discretization errors also 
plays a critical role in the Lax-Wendroff method because it allows the second 
derivative of the solution in time to be written in terms of the second 
derivative in space.  This observation transforms an unstable, first-order 
scheme based on a central difference approximation of 
$\partial/\partial x$ into a stable, second-order 
scheme~\cite{leveque_book_1992, leveque_book_2002}.

Optimal time step selection as a technique for improving the overall accuracy 
of numerical schemes for time dependent PDEs appears to be a novel idea.
Typically, the primary concern when choosing a time step is numerical 
stability with accuracy being a secondary consideration.  The idea that a 
clever choice of time step can be used to dramatically reduce the error is 
rarely considered.  In general, little research seems to have been done in
the area of optimizing the parameters of numerical methods for PDEs (\eg time 
step and grid spacing) to improve accuracy.  To the best of the author's 
knowledge, Zhao's method for choosing an optimal value of $\theta$ in the 
theta-method is the only work that attempts to improve accuracy by modifying 
numerical parameters.  While he focuses solely on the optimal choice of 
$\theta$ for a fixed time step, his approach can be viewed as a form of 
optimal time step selection by reversing the roles of $\theta$ and the time 
step.

It is worth spending a few moments to draw the distinction between optimal 
time step selection and traditional adaptive time stepping 
techniques~\cite{iserles_book, shampine_2005} that are used in the 
context of the method 
of lines~\cite{iserles_book, gko_book}.  On the surface, optimal time step 
selection may appear to be just a crude version of adaptive time step control.
However, the two methods are actually very different.  While both methods seek 
to reduce numerical error by controlling the time step, the numerical errors 
they control are fundamentally different.  Adaptive time stepping can only 
be used to reduce temporal discretization errors because it controls the 
errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Thus, optimal time 
step selection can potentially produce superior results compared with 
traditional adaptive time stepping approaches for integrating the 
semi-discrete equations associated with time dependent PDEs.  Another 
important distinction between the two techniques is that optimal time step 
selection uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

@@

The remainder of this paper is organized as follows.  
In Section~\ref{sec:optimal_time_step_selection}, we begin by 
discussing the technique of optimal time step selection and 
presenting the analysis that explains how it boosts the order of accuracy 
for formally low-order numerical schemes.  In 
Section~\ref{sec:applications_1d}, we demonstrate the general utility of 
optimal time selection to problems in one space dimension by applying it to 
finite difference schemes for several classical linear and nonlinear PDEs. 
In Section~\ref{sec:applications_multidim}, we use optimal time step selection 
to improve the accuracy of simple finite difference schemes for the linear 
advection equation and diffusion equation in multiple space dimensions.  
To illustrate the ease with which high-order accuracy can be achieved for
problems on arbitrary domains, the 2D diffusion equation is solved on both 
the regular and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size and order of accuracy.  
Finally, in Section~\ref{sec:summary}, we provide some concluding remarks.

\section{\label{sec:optimal_time_step_selection} 
         Optimal Time Step (OTS) Selection}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather it is a way to enhance to the performance
of existing finite difference schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  The two basic ideas underlying 
optimal time step selection are (1) the PDE provides valuable insight into the 
discretization errors for finite difference schemes and (2) a judicious choice 
of time step can be used to eliminate the leading-order (and possibly 
higher-order) terms in the error.  In this section, we present and analyze 
the technique of optimal time step selection for finite difference schemes 
constructed to solve evolution equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important in 
applied mathematics, restricting our attention to problems of this form is 
not a serious limitation.

@@

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the finite difference approximation of the spatial 
derivative operator acting on ${\mathbf u}$.  The second term in the local 
error (\ref{eq:leading_order_error_model_PDE_MOL}) arises directly from 
temporal discretization errors associated with the numerical scheme used to 
integrate the system of ODEs in time.  The first term arises because $F$ is 
only an approximation of the spatial derivative operators in the PDE.  As a 
result, the system of ODEs~(\ref{eq:method_of_lines}) has an error due to 
the spatial discretization at \emph{every} point in time.  For a spatial 
discretization error of order $r$, the error in (\ref{eq:method_of_lines}) is 
$O(\dx^r)$.  Therefore, the error accumulated during each time step due to 
spatial discretization error is $O(\dt \dx^r)$ regardless of how the system of 
ODEs (\ref{eq:method_of_lines}) is solved numerically.  Because the method of 
lines is so frequently used to construct finite difference approximations for 
time dependent PDEs, we will focus solely on finite difference schemes of this 
type for the remainder of the paper to keep the discussion simple.  It is 
straightforward to extend the results we obtain to more general finite 
difference schemes.

To analyze the accuracy of the finite difference scheme 
for~(\ref{eq:model_PDE}) when the optimal time step is used, we examine the 
higher-order terms in the discretization error following the conventional 
approach (\ie ``big-O'' notation).  Elimination of the leading terms in the 
discretization error leaves a local error of the form 
$O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ and $(q+1) > (s+1)$ are the orders 
of the spatial and temporal discretization errors, respectively, after the 
leading-order terms in the error have been eliminated.  As for the 
leading-order error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of 
the local error is a direct consequence of the method of lines approach to 
discretizing the PDE.

@@

@@

% TABLE SHOWING COMPUTATIONAL COSTS DISCUSSED IN SECTION
% sec:computational_performance.  PLACED HERE TO SPREAD TABLES OUT
\begin{table}[!b]
\caption{\label{tab:comp_perf_vs_err} 
   Computational cost of various finite difference schemes for the 1D 
   diffusion equation as a function of the global numerical error $e$.
   For all of the schemes, the standard second-order central difference 
   stencil is used to discretize the Laplacian, and the numerical 
   solution is assumed to be computed over a fixed time interval.  
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|l|c|c|c|c|}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$\footnote{The 
   time step for the forward/backward Euler and Crank-Nicholson schemes 
   are $O(\dx^2)$ and $O(\dx)$, respectively.  Note that the 
   time step for the backward Euler scheme is chosen so that the temporal 
   error is subdominant to the spatial error.}
  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
    to be $O(1/\dx)$.} 
  & {\bf Compute Time}\footnote{The computation time is approximated as 
    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
    arise for the implicit methods.}  \\
  \hline 
  Forward Euler    & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Backward Euler   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-1} \right)$ \\
  Forward Euler with OTS  & $O\left( e^{1/4} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/4} \right)$ 
                   & $O\left( e^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{center}
\end{minipage}
\end{table}

@@

\begin{table}[!t]
\caption{\label{tab:comp_perf_vs_dim}
   Computational cost of the forward Euler scheme with optimal time step and 
   the Crank-Nicholson method for solving the diffusion equation as a function 
   of the global numerical error $e$.
   For both schemes, second-order, isotropic stencils are used to discretize 
   the Laplacian.  For a problem in $d$ space dimensions, the memory 
   requirement and computation time for each scheme are estimated to be 
   $O \left(1/\dx^d \right)$ and 
   $O\left( \left(1/\dx^d \right) (1/\dt) \right)$, respectively, which 
   assumes that we are using an efficient 
   $O(N) = O \left( 1/\dx^d \right)$ algorithm to invert the matrices that 
   arise for the Crank-Nicholson method.
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|c|c|}
  \hline
  & \multicolumn{2}{|c|}{{\bf Forward Euler with OTS}\footnote{For the forward 
    Euler scheme, the optimal time step is $\dto = \dx^2/6$.}}
  & \multicolumn{2}{|c|}{\bf Crank-Nicholson} \\
  \cline{2-3} \cline{4-5} 
    {\bf Dimensions} & {\bf Memory} & {\bf Compute Time} 
  & {\bf Memory} & {\bf Compute Time} \\
  \hline 
  $1$ & $O\left( e^{-1/4} \right)$ 
      & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ \\ 
  $2$ & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-3/2} \right)$ \\ 
  $3$ & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-5/4} \right)$ 
      & $O\left( e^{-3/2} \right)$ 
      & $O\left( e^{-2} \right)$ \\
  $d$ & $O\left( e^{-d/4} \right)$ 
      & $O\left( e^{-(d+2)/4} \right)$ 
      & $O\left( e^{-d/2} \right)$ 
      & $O\left( e^{-(d+1)/2} \right)$ \\ 
  \hline 
\end{tabular}
\end{center}
\end{minipage}
\end{table}

@@

\subsection{Boundary Conditions}
In our analysis and discussion thus far, we have completely ignored the issue 
of boundary conditions.  The choice of discretization for boundary conditions
can have a significant impact on the accuracy of the numerical solution.  In 
general, high-order discretizations of boundary conditions are required to 
ensure that errors introduced at the domain boundary do not pollute the 
numerical solution in the domain interior.  For Dirichlet boundaries, it
is sufficient to impose boundary conditions by using boundary values that have 
the same order of accuracy as the solution.  More effort is required to ensure 
that Neumann boundary conditions do not reduce the order of accuracy of the 
solution.  In general, Neumann boundary conditions require that the accuracy 
of ghost cells slightly exceed the desired order of accuracy for the 
solution.  A detailed discussion of methods for imposing boundary conditions 
is beyond the scope of this paper.  We refer interested readers to the 
thorough analysis presented in~\cite{gko_book}.  To avoid complicating the 
discussion in Sections~\ref{sec:applications_1d} 
and~\ref{sec:applications_multidim}, Neumann boundary conditions are imposed 
by taking advantage of knowledge of the analytical solution.
